Relative parabolic factorizations of Chevalley groups

### 2015-04-11 dennis-vaserstein.tex

\begin{lemma}\label{Stein_reduction} 
Suppose that there exists a subset $\widetilde{L_r} \subseteq \E(\Delta_r, R, I)$ satisfying the following two conditions:
\begin{enumerate}[label=(\alph*)] 
 \item\label{stein_cond1} One has $\U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot \widetilde{L_r} = \U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot \E(\Delta_r, R, I).$
 \item\label{stein_cond2} There is an inclusion $X_{-\alpha\ssub{r}}(I)^{\widetilde{L_r}} \subseteq \EP_s(R, I).$
\end{enumerate}
Then $A_{rs} = \E(\Phi, R, I)$. \end{lemma}
\begin{proof} Set $A:=\U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot \widetilde{L_r} \cdot \EP_s(\Phi, R, I).$
From condition~\ref{stein_cond1} of the lemma it follows that $A_{rs}=A$.

@@

Now we are ready to prove the needed equality. By virtue of Theorem~\ref{theorem:Stepanov} it suffices to show that
$A$ is invariant with respect to left multiplication by elements of the following two families:
\begin{itemize}
 \item $z_{\alpha}(s, \xi)$, $s\in I$, $\xi \in R$, $\alpha\in\Sigma^-_s$;
 \item $x_{\beta}(s)$, $s \in I$, $\beta \in \Phi$. \end{itemize}
In fact, thanks to Lemma~\ref{lemma:conj_lemma} it suffices to prove this statement only for the second family.
Moreover, we can additionally assume $\beta\in \Sigma_r^-$.
%TODO: To be continued 
\end{proof}

>>>>

\begin{lemma}\label{Stein_reduction}
Let $\Phi_l = \rA_l, \rD_l, \rE_6, \rE_7$. If $\Phi\neq \rE_7$ set $s:=1$ otherwise set $s:=7$. 
Let $r$ be any other natural number not equal to $s$.
Assume that there exists a subset $\widetilde{L_r} \subseteq \E(\Delta_r, R, I)$ satisfying the following two conditions:
\begin{enumerate}[label=(\alph*)] 
 \item\label{stein_cond1} One has $\U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot \widetilde{L_r} = \U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot \E(\Delta_r, R, I).$
 \item\label{stein_cond2} There is an inclusion $X_{-\alpha\ssub{r}}(I)^{\widetilde{L_r}} \subseteq \EP_s(R, I).$
\end{enumerate}
Then $A_{rs} = \E(\Phi, R, I)$. \end{lemma}
\begin{proof} Set $A:=\U(\Phi^+, I)\cdot \U(\Phi^-, I) \cdot \widetilde{L_r} \cdot \EP_s(\Phi, R, I).$
From condition~\ref{stein_cond1} of the lemma it follows that $A_{rs}=A$.

@@

Now we are ready to prove the needed equality. By virtue of Theorem~\ref{theorem:Stepanov} it suffices to show that
$A$ is invariant with respect to left multiplication by elements of the following two families:
\begin{itemize} \item $z_{\alpha}(s, \xi)$, $s\in I$, $\xi \in R$, $\alpha\in\Sigma^-_s$;
 \item $x_{\beta}(s)$, $s \in I$, $\beta \in \Phi$. \end{itemize}
Thanks to Lemma~\ref{lemma:conj_lemma} it suffices to check the required statement only for the second family of elements.
Denote by $S$ the set of all roots $\alpha$ of $\Phi$ for which one has $X_{\alpha}(I)\cdot A_{rs}\subseteq A_{rs}$.
It remains to prove that $S$ coincides with $\Phi$. We already know that $\Phi^+ \subseteq S$.

Notice that the Weyl group $W(\Delta_s)$ acts transitively on $\Delta_s$, therefore in view of~\ref{rel:R3} the subgroup
$W_s := \langle w_\alpha(1) \mid \alpha\in\Delta_s\rangle \leq \EP_s$ acts transitively on the set of root subgroups $X_\alpha(I)$, $\alpha\in \Delta_s$.
Since $-\alpha_r\in S$ this implies that that $\Delta_s\subseteq S$.

Now denote by $\widetilde{\alpha}$ the maximal root of $\Phi$. Our assumptions on $\Phi$ guarantee that $m_s(\alpha)=1$, 
hence every two roots $\alpha, \beta \in \Sigma^-_s$ have the same $s$-shape (i.\,e. $\shape(\{s\}, \alpha = \shape(\{s\}, \beta)$).
By Lemma~\ref{lemma:abs} $W(\Delta_s)$ interchanges $\alpha$ and $\beta$. %if their length is equal.
Since $\alpha_s \in S$ the argument similar to the one above implies $\Sigma^-_s\subseteq S$. This completes the proof of the lemma. \end{proof}

### 2015-04-11 paper.tex

\DeclareMathOperator{\K}{K}
\DeclareMathOperator{\G}{G}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\St}{St}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\EP}{EP}
\DeclareMathOperator{\U}{U}
\DeclareMathOperator{\X}{X}
\DeclareMathOperator{\Z}{Z}
\DeclareMathOperator{\M}{M}
\DeclareMathOperator{\SR}{SR}
\DeclareMathOperator{\sr}{sr}
\DeclareMathOperator{\Rad}{Rad}
\DeclareMathOperator{\Max}{Max}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\ASR}{ASR}
\DeclareMathOperator{\asr}{asr}
%TODO: Deprecate Um -- Use Ums/Umd whenever possible
\DeclareMathOperator{\Um}{Um}
\DeclareMathOperator{\Ums}{Ums}
\DeclareMathOperator{\Umd}{Umd}
\DeclareMathOperator{\rk}{rk}
\newcommand{\rA}{\mathsf{A}}
\newcommand{\rB}{\mathsf{B}}
\newcommand{\rC}{\mathsf{C}} 
\newcommand{\rD}{\mathsf{D}} 
\newcommand{\rE}{\mathsf{E}}

@@

The following theorem is the main result of \cite{S}.
\begin{thm}\label{theorem:Stepanov}
Let $\Phi$ be a reduced irreducible root system and let $S$ be any parabolic subset of roots of $\Phi$ with special part $\Sigma_S$.
Then $\E(\Phi, R, I)$ is generated by the following two families of elements:
\begin{itemize}
 \item $x_{\alpha}(s)$, where $s\in I$, $\alpha\in\Phi$;
 \item $z_\alpha(s,\xi)$, where $s\in I$, $\xi\in R$, $\alpha\in\Sigma_S$. \end{itemize} \end{thm}
\begin{proof} See \cite[Theorem~3.4]{S}. \end{proof}

>>>>

\DeclareMathOperator{\K}{K}
\DeclareMathOperator{\G}{G}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\St}{St}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\EP}{EP}
\DeclareMathOperator{\U}{U}
\DeclareMathOperator{\X}{X}
\DeclareMathOperator{\Z}{Z}
\DeclareMathOperator{\M}{M}
\DeclareMathOperator{\SR}{SR}
\DeclareMathOperator{\sr}{sr}
\DeclareMathOperator{\shape}{shape}
\DeclareMathOperator{\Rad}{Rad}
\DeclareMathOperator{\Max}{Max}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\ASR}{ASR}
\DeclareMathOperator{\asr}{asr}
%TODO: Deprecate Um -- Use Ums/Umd whenever possible
\DeclareMathOperator{\Um}{Um}
\DeclareMathOperator{\Ums}{Ums}
\DeclareMathOperator{\Umd}{Umd}
\DeclareMathOperator{\rk}{rk}
\newcommand{\rA}{\mathsf{A}}
\newcommand{\rB}{\mathsf{B}}
\newcommand{\rC}{\mathsf{C}} 
\newcommand{\rD}{\mathsf{D}} 
\newcommand{\rE}{\mathsf{E}}

@@

\subsection{Chevalley groups}
Our exposition of Chevalley groups is standard and follows (...).
\textbf{TODO: Write more about Chevalley groups}

Recall that elementary root unipotents $t_{\alpha}(\xi)$, $x\in R$, $\alpha\in\Phi$ satisfy the following set of relations called Steinberg relations.
\begin{equation}\label{rel:add} t_\alpha(s) t_\alpha(t) = t_\alpha(s+t), \end{equation} 
\begin{equation}\label{rel:CCF} [t_\alpha(s),  t_\beta(t)] = \left(\prod\limits t_{i\alpha + j\beta}(N_{\alpha,\beta, i, j}s^i t^j)\right),\quad \alpha\neq-\beta, N_{\alpha, \beta, i, j}\in\mathbb{Z}.\end{equation}
The indices $i$, $j$ in the above formula range over all positive natural numbers such that $i\alpha + j\beta\in\Phi$.
The integers $N_{\alpha, \beta, i, j}$ are called {\it structure constants} of the Chevalley group $\G(\Phi,R)$ and depend only on $\Phi$.

For $\xi\in R^*$ set $w_\alpha(\xi) := x_\alpha(\xi) x_{-\alpha}(-\xi^{-1}) x_{\alpha}(\xi).$
If $\rk(\Phi)\geq 2$ the following relation is a consequence of \ref{rel:add}--\ref{rel:CCF}:
\begin{equation}\label{rel:R3} w_\alpha(\xi) t_{\beta}(s) w_\alpha(\xi)^{-1} =
t_{\sigma\ssub{\alpha}\beta} (\eta_{\alpha, \beta}\xi^{-\langle\beta,\alpha \rangle} s),\ \xi,s\in R. \end{equation}

Let $\Pi=\{\alpha_1, \ldots, \alpha_l\}$ be a basis of $\Phi$. Denote by $\Phi^+$ and $\Phi^-$ subsets of positive and negative roots with respect to $\Pi$.
Denote by $m_i(\alpha)$ the coefficient of $\beta$ in the expansion of $\alpha$ in $\Pi$, i.\,e. $\alpha = \sum\limits_{i=1}^n m_i(\alpha) \alpha_i$.

A root subset $S\subseteq \Phi$ is called {\it parabolic} (resp. {\it reductive}, resp. {\it special}) if $\Phi=S \cup -S$ (resp. $S = -S$, resp. $S \cap -S=\varnothing$).
Any parabolic subset $S \subseteq \Phi$ can be decomposed into the disjoint union of its reductive and special part, i.e. 
$S = \Sigma_S \sqcup \Delta_S$, where $\Sigma_S \cap (-\Sigma_S) = \varnothing$, $\Delta_S = -\Delta_S$.

We denote by $(\alpha, \beta)$ the scalar product of roots and by $\langle \alpha, \beta\rangle$ the integer $2(\alpha, \beta)/(\alpha, \alpha)$.
Denote by $W(\Phi)$ the subgroup of isometries of $\Phi$ generated by all reflections $\sigma_\alpha$ where $\sigma_\alpha(\beta)=\beta-\langle\alpha,\beta \rangle\cdot \alpha$.
For $S\subseteq \Phi$ denote by $W(S)$ the subgroup of $W(\Phi)$ generated by $\sigma_\alpha$ for $\alpha\in S$.

Let $J\subseteq \Pi$ be a subset of simple roots. 
For a root $\beta = \sum d_i\alpha_i \in \Phi$ set $\shape(J, \beta):=\sum\limits_{i\in J}d_i \alpha_i$.
Set $$\Delta_J = \{\alpha \in \Phi \mid \shape(J, \alpha)=0\},\quad \Sigma^\pm_J = \{\alpha \in \Phi \mid \shape(J, \alpha)\in\Phi^\pm \}.$$
Clearly, $\Delta_J$ is a reductive subset of roots while $\Sigma^\pm_J$ are special subsets.

\begin{lemma}\label{lemma:abs} Let $\alpha, \beta \in \Sigma^\pm_J$ be two roots having the same length such that $\shape(J,\alpha)=\shape(J,\beta)\neq 0$.
Then $\alpha$ and $\beta$ are conjugate under the action of $W(\Delta_J)$. \end{lemma} \begin{proof} See~\cite[Lemma~1]{ABS}. \end{proof}

\subsection{Relative elementary subgroup}
The following theorem is the main result of \cite{S}.
\begin{thm}\label{theorem:Stepanov}
Let $\Phi$ be a reduced irreducible root system and let $S$ be any parabolic subset of roots of $\Phi$ with special part $\Sigma_S$.
Then $\E(\Phi, R, I)$ is generated by the following two families of elements:
\begin{itemize}
 \item $x_{\alpha}(s)$, where $s\in I$, $\alpha\in\Phi$;
 \item $z_\alpha(s,\xi)$, where $s\in I$, $\xi\in R$, $\alpha\in\Sigma_S$. \end{itemize} \end{thm}
\begin{proof} See~\cite[Theorem~3.4]{S}. \end{proof}
